Weighted Model Counting with Conditional Weights for Bayesian Networks

### 2021-02-08 doc/paper2/paper.tex
Updated plots
Fonts, references, etc.

\DeclareMathOperator{\im}{im}

@@

\begin{abstract}
  Weighted model counting (WMC) has emerged as the unifying inference mechanism
  across many (probabilistic) domains. Encoding an inference problem as an
  instance of WMC typically necessitates adding extra literals and clauses. This
  is partly so because the predominant definition of WMC assigns weights to
  models based on weights on literals, and this severely restricts what
  probability distributions can be represented. We develop a measure-theoretic
  perspective on WMC and propose a way to encode conditional weights on literals
  analogously to conditional probabilities. This representation can be as
  succinct as standard WMC with weights on literals but can also expand as
  needed to represent probability distributions with less structure. To
  demonstrate the performance benefits of conditional weights over the addition
  of extra literals, we develop a new WMC encoding for Bayesian networks and
  adapt a state-of-the-art WMC algorithm ADDMC to the new format. Our
  experiments show that the new encoding results in significantly faster
  inference on most benchmark instances. Furthermore, the same idea can be
  adapted to other WMC algorithms and other problem domains.
\end{abstract}

@@

@@

This new measure-theoretic perspective allows us to not only encode any
discrete probability distribution but also improve inference speed. In
\cref{sec:bns}, we demonstrate this by developing a new WMC encoding for
Bayesian networks that uses \emph{conditional weights} on literals (in the
spirit of conditional probabilities) that have literal-based WMC as a special
case. We prove the correctness of the encoding and show how a state-of-the-art
WMC solver ADDMC \citep{DBLP:conf/aaai/DudekPV20} can be adapted to the new
format. ADDMC is a recently-proposed algorithm for WMC based on manipulating
functions on Boolean algebras using an efficient representation for such
functions known as algebraic decision diagrams (ADDs)
\citep{DBLP:journals/fmsd/BaharFGHMPS97}. Finally, in \cref{sec:experiments}, we
show how the new encoding results in faster inference on the vast majority of
benchmark instances, often by one or two orders of magnitude. We explain the
performance benefits by showing how our encoding has asymptotically fewer
variables and ADDs.

@@

Hitherto, four techniques have been proposed for encoding Bayesian networks into
instances of WMC. We will identify them based on the initials of authors as well
as publications years: \texttt{d02} \citep{DBLP:conf/kr/Darwiche02},
\texttt{sbk05} \citep{DBLP:conf/aaai/SangBK05}, \texttt{cd05}
\citep{DBLP:conf/ijcai/ChaviraD05}, and \texttt{cd06}
\citep{DBLP:conf/sat/ChaviraD06}. Below we summarise the observed performance
differences among them. \citet{DBLP:conf/aaai/SangBK05} claim that
\texttt{sbk05} is a smaller encoding than \texttt{d02} with respect to both the
number of clauses and the number of variables but provide no experimental
comparison. \citet{DBLP:conf/ijcai/ChaviraD05} compare \texttt{cd05} with
\texttt{d02} by measuring the time it takes to compile either encoding into an
arithmetic circuit (but do not measure inference time). The more recent encoding
\texttt{cd05} always compiles faster and results in a smaller arithmetic circuit
(as measured by the number of edges). In their subsequent paper, the same
authors perform two sets of experiments (that are relevant to this summary)
\citep{DBLP:conf/sat/ChaviraD06}. First, they compile \texttt{cd05} and
\texttt{cd06} encodings into d-DNNF (i.e., deterministic decomposable negation
normal form \citep{DBLP:journals/jancl/Darwiche01}), measuring both compilation
time and numbers of edges in the d-DNNF diagram. The results are mostly in
favour of \texttt{cd06}. Second, they compare the inference time of
\texttt{sbk05} run with Cachet \citep{DBLP:conf/sat/SangBBKP04} with the compile
times of \texttt{cd05} and \texttt{cd06}, but only on five (types of) instances.
In these experiments, \texttt{cd06} is always faster than \texttt{cd05}, while
the comparison with \texttt{sbk05} is mixed. Sometimes \texttt{cd06} is orders
of magnitude faster than \texttt{sbk05}, sometimes slightly slower. The
performance difference between \texttt{sbk05} and \texttt{cd05} is even harder
to judge: \texttt{sbk05} is better on three out of five instances and worse on
the remaining two. Based on this description, one would expect \texttt{cd06} to
be faster than both \texttt{cd05} and \texttt{sbk05}, both of which should be
faster than \texttt{d02}. The experiments in \cref{sec:experiments}, however,
strongly disagree with this prediction, showing that the quality of an encoding
depends strongly on the underlying search algorithm or compilation technique.

% \paragraph{ADDs and their use in inference.} ADDs provide an efficient way to
% manipulate functions from a Boolean algebra to any algebraic structure (most
% commonly real numbers) \citep{DBLP:journals/fmsd/BaharFGHMPS97}. They have been
% used to represent value functions in Markov decision processes (MDPs)
% \citep{DBLP:conf/uai/HoeySHB99} and extended to handle first-order MDPs
% \citep{DBLP:journals/ai/SannerB09} and discrete and continuous state MDPs
% \citep{DBLP:conf/uai/SannerDB11}. The latter extension has also been used for
% weighted model integration---an extension of WMC that can handle continuous
% variables \citep{DBLP:conf/ijcai/KolbMSBK18}. For Bayesian networks, ADDS have
% been used to represent each conditional probability table (CPT) as an ADD
% \citep{DBLP:conf/icml/ZhaoMP15} and as a combination of ADDs and arithmetic
% circuits \citep{DBLP:conf/ijcai/ChaviraD07}. ADDs are particularly advantageous
% in this situation because of their ability to fully exploit context-specific
% independence, i.e., the observable structure within a CPT that is not inherited
% from the structure of the Bayesian network \citep{DBLP:conf/uai/BoutilierFGK96}.

@@

We also consider the space of all functions from any Boolean algebra to
$\mathbb{R}_{\ge 0}$ together with some operations on those functions. They will
be instrumental in defining WMC as a measure in \cref{sec:wmc_as_measure} and
can be efficiently represented using ADDs. Furthermore, all of the operations
are supported by the CUDD \citep{somenzi1998cudd} package that is used by ADDMC
for ADD manipulation \citep{DBLP:conf/aaai/DudekPV20}. The definitions of
multiplication and projection are as defined by
\citet{DBLP:conf/aaai/DudekPV20}, while others are new.

@@

\Cref{alg:encoding} produces a function with a Boolean algebra as its domain.
This function can be represented by an ADD
\citep{DBLP:journals/fmsd/BaharFGHMPS97}. The core of ADDMC works by taking an
ADD $\psi\colon 2^{U} \to \mathbb{R}_{\ge 0}$ (expressed as a product of smaller
ADDs) and returning $(\exists_U\psi)(\emptyset)$
\citep{DBLP:conf/aaai/DudekPV20}. In this section, we prove that the function
$\phi$ produced by \cref{alg:encoding} can be used by ADDMC to correctly compute
any marginal probability of the Bayesian network that was encoded as
$\phi$.\footnote{Note that it can just as well compute \emph{any} probability
  expressed using the random variables in $\mathcal{V}$.} We begin with
\cref{lemma:cpt} which shows that any conditional weight function produces the
right answer when given a valid encoding of variable-value assignments relevant
to the CPT.

@@

We end with \cref{thm:correctness} that shows how $\phi$ can be combined with an
encoding of a single variable-value assignment so that ADDMC would compute its
marginal probability.

@@

\Cref{alg:encoding} encodes a Bayesian network into a function on a Boolean
algebra, but how does it relate to the standard interpretation of a WMC encoding
as a formula in conjunctive normal form (CNF) together with a collection of
weights? The factors of $\phi$ that restrict the values of indicator variables
for non-binary random variables are already expressed as a product of sums of
0/1-valued functions, i.e., a kind of CNF. Disregarding these functions, each
conditional weight function $\mathrm{CPT}_X$ is represented by a sum with a term
for every subset of $\mathcal{E}^*(X)$. To encode these terms, we introduce
\emph{extended weight clauses} to the WMC format used by Cachet
\citep{DBLP:conf/sat/SangBBKP04}. For instance, here is a representation of the
Bayesian network from \cref{fig:example_bn}:
\[
  \begin{array}{lrrll}
    \lambda\sb{T=l} &\lambda\sb{T=m} &\lambda\sb{T=h} & &0 \\
                    &-\lambda\sb{T=l} &-\lambda\sb{T=m} & &0 \\
                    &-\lambda\sb{T=l} &-\lambda\sb{T=h} & &0 \\
                    &-\lambda\sb{T=m} &-\lambda\sb{T=h} & &0 \\
    w &\lambda\sb{W=1} & &0.5 &0.5 \\
    w &\lambda\sb{F=1} &\lambda\sb{W=1} &0.6 &0.4 \\
    w &\lambda\sb{F=1} &-\lambda\sb{W=1} &0.1 &0.9 \\
    w &\lambda\sb{T=l} &\lambda\sb{W=1} &0.2 &1 \\
    w &\lambda\sb{T=m} &\lambda\sb{W=1} &0.4 &1 \\
    w &\lambda\sb{T=h} &\lambda\sb{W=1} &0.4 &1 \\
    w &\lambda\sb{T=l} &-\lambda\sb{W=1} &0.6 &1 \\
    w &\lambda\sb{T=m} &-\lambda\sb{W=1} &0.3 &1 \\
    w &\lambda\sb{T=h} &-\lambda\sb{W=1} &0.1 &1
  \end{array}
\]
where each indicator variable is eventually replaced with a unique positive
integer. Each line prefixed with a $w$ can be split into four parts: the `main'
variable (always not negated), conditions (possibly none), and two weights. For
example, the line
\[
  \begin{array}{lrrll}
    w &\lambda\sb{T=m} &-\lambda\sb{W=1} &0.3 &1
  \end{array}
\]
encodes the function $0.3[\lambda_{T=m}] \cdot \overline{[\lambda_{W=1}]} +
1\overline{[\lambda_{T=m}]} \cdot \overline{[\lambda_{W=1}]}$ and can be
interpreted as defining two conditional weights: $\nu(T = m \mid W = 0) = 0.3$,
and $\nu(T \ne m \mid W = 0) = 1$, the former of which corresponds to a row in
the CPT of $T$ while the latter is artificially added as part of the encoding.
In our encoding of Bayesian networks, it is always the case that, in each weight
clause, either both weights sum to one, or the second weight is equal to one.
Finally, note that the measure induced by these weights is not probabilistic
(i.e., $\mu(\top) \ne 1$) by itself, but it becomes probabilistic when combined
with the additional clauses that restrict what combinations of indicator
variables can co-occur.

\subsection{Changes to ADDMC}

Here we describe two changes to
ADDMC\footnote{\url{https://github.com/vardigroup/ADDMC}}
\citep{DBLP:conf/aaai/DudekPV20} needed to adapt it to the new format.

First, ADDMC constructs the \emph{Gaifman graph} \citep{gaifman1982local} of the
input CNF formula as an aid for the algorithm's heuristics. This graph has as
vertices the variables of the formula, and there is an edge between two
variables $u$ and $v$ if there is a clause in the formula that contains both $u$
and $v$. We extend this definition to functions on Boolean algebras, i.e., the
factors of $\phi$. For any pair of distinct variables $u, v \in U$, we draw an
edge between them in the Gaifman graph if there is a function $\alpha\colon 2^X
\to \mathbb{R}_{\ge 0}$ that is a factor of $\phi$ such that $u \in X$ and $v
\in X$. For instance, a factor such as $\mathrm{CPT}_X$ will enable edges
between all distinct pairs of variables in $\mathcal{E}^*(X)$.

Second, even though the function $\phi$ produced by \cref{alg:encoding} is
constructed to have $2^U$ as its domain, sometimes the domain is effectively
reduced to $2^V$ for some $V \subset U$ by the ADD manipulation algorithms that
optimise the ADD representation of a function. For a simple example, consider
$\alpha: 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ defined as $\alpha(\{a\}) =
\alpha(\emptyset) = 0.5$. Then $\alpha$ can be reduced to $\alpha'\colon
2^{\emptyset} \to \mathbb{R}_{\ge 0}$ defined as $\alpha'(\emptyset) = 0.5$. To
compensate for these reductions, for the original WMC format with a weight
function $w\colon U \cup \{ \neg u \mid u \in U \} \to \mathbb{R}_{\ge 0}$,
ADDMC would multiply its computed answer by $\prod_{u \in U \setminus V} w(u) +
w(\neg u)$. With the new WMC format, we instead multiply the answer by $2^{|U
  \setminus V|}$. Each `excluded' variable $u \in U \setminus V$ satisfies two
properties: all weights associated with $u$ are equal to $0.5$ (otherwise the
corresponding CPT would depend on $u$, and $u$ would not be excluded), and all
other CPTs are independent of $u$ (or they may have a trivial dependence, where
the probability stays the same if $u$ is replaced with its complement). Thus,
the CPT that corresponds to $u$ still multiplies the weight of every atom in the
Boolean algebra by $0.5$, but the number of atoms under consideration is halved.
To correct for this, we multiply the final answer by two for every $u \in U
\setminus V$.

@@

In this section, we describe an experimental study comparing the five WMC
encodings for Bayesian networks when run with ADDMC. The experiments were run on
servers with Intel Xeon Gold 6138 and Intel Xeon E5-2630
processors\footnote{Each instance is run on the same processor for all
  encodings.} running Scientific~Linux~7 with an \SI{8}{\gibi\byte} memory limit
and a \SI{1000}{\second} timeout.

@@

For all encodings other than \texttt{cw}, we use their implementation in
Ace~3.0\footnote{\url{http://reasoning.cs.ucla.edu/ace/}} with
\texttt{-encodeOnly} and \texttt{-noEclause} flags. However, Ace was not used to
encode evidence, as preliminary experiments revealed that the evidence-encoding
implementation contains bugs that can lead to incorrect answers or a Java
exception being thrown on some instances of the data set (and the source code is
not publicly available). Instead, we simply list all the evidence as additional
clauses in the encoding.

Note that both \texttt{cd05} and \texttt{cd06} purposefully produce overly
relaxed encodings that contain extra models and thus yield incorrect
probabilities \citep{DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06}. These
additional models are supposed to be filtered out during circuit compilation
\citep{DBLP:conf/ijcai/ChaviraD05}, but this is not easily achievable with
ADDMC. Nonetheless, we include both encodings in the experiments.

@@

For experimental data, we use the Bayesian networks available with Ace and
Cachet\footnote{\url{https://www.cs.rochester.edu/u/kautz/Cachet/}}, most of
which happen to be binary. We classify them into the following seven categories:
\begin{itemize*}
\item DQMR and
\item Grid networks as described by \citet{DBLP:conf/aaai/SangBK05},
\item Mastermind, and
\item Random Blocks from the work of \citet{DBLP:journals/ijar/ChaviraDJ06},
\item remaining binary Bayesian networks that include Plan Recognition
  \citep{DBLP:conf/aaai/SangBK05}, Friends and Smokers, Students and Professors
  \citep{DBLP:journals/ijar/ChaviraDJ06}, and \texttt{tcc4f}, and
\item non-binary classic Bayesian networks (\texttt{alarm}, \texttt{diabetes},
  \texttt{hailfinder}, \texttt{mildew}, \texttt{munin1}--\texttt{4},
  \texttt{pathfinder}, \texttt{pigs}, \texttt{water}).
\end{itemize*}
We run ADDMC with each of the five encodings once on each Bayesian network.

\begin{figure}
  \centering
  \input{cumulative.tex}%
  \caption{Cumulative number of instances solved by ADDMC over time using each
    encoding.}
  \label{fig:cumulative}
\end{figure}

\begin{table}
  \centering
  \caption{The numbers of instances (out of 1466) solved by ADDMC with each
    encoding (uniquely, faster than with others, and in total) under the
    described constraints.}%
  \begin{tabular}{lrrr}
    \toprule
    Encoding & Unique & Fastest & Total \\
    \midrule
    \texttt{cd05} & 2 & 3 & 380 \\
    \texttt{cd06} & 0 & 0 & 363 \\
    \texttt{cw} & 234 & 935 & 980 \\
    \texttt{d02} & 40 & 111 & 797 \\
    \texttt{sbk05} & 12 & 44 & 765 \\
    \bottomrule
  \end{tabular}
  \label{tbl:tallies}
\end{table}

\begin{figure*}
  \centering
  \input{scatter.tex}%
  \caption{ADDMC inference time using \texttt{cw} compared to \texttt{d02}
    (left) and \texttt{sbk05} (right) for each instance in all data sets.}
  \label{fig:scatter}
\end{figure*}

According to the cumulative plot in \cref{fig:cumulative}, \texttt{cw} is
consistently able to solve more instances with any given time limit than any
other encoding, while the order of the remaining encodings from best to worst is
exactly the opposite of that in previous literature (cf. \cref{sec:related}).
Specifically, \texttt{cw} solves the same number of instances in
\SI{8.144}{\second} as \texttt{d02} does in \SI{1000}{\second}, which is the
second-best encoding in our experiments. Furthermore, \cref{tbl:tallies} shows
that \texttt{cw} was the fastest in 935 out of the 980 (i.e.,
\SI{95.4}{\percent}) instances that it managed to solve. Finally, according to
the scatter plots in \cref{fig:scatter}, \texttt{cw} is particularly promising
on Grid networks while losing to \texttt{d02} on some instances of the
Mastermind data set. We conjecture that this is so either because the particular
structure of the problem confuses the heuristics used by ADDMC or because these
instances allow for significant simplifications that are exploited by
\texttt{d02} but not \texttt{cw}.

@@

\newpage
\bibliography{paper}

\newpage
\appendix
\section{Proofs}

>>>>

\DeclareMathOperator{\im}{im}
\Crefname{algocf}{Algorithm}{Algorithms}

@@

\begin{abstract}
  Weighted model counting (WMC) has emerged as the unifying inference mechanism
  across many (probabilistic) domains. Encoding an inference problem as an
  instance of WMC typically necessitates adding extra literals and clauses. This
  is partly so because the predominant definition of WMC assigns weights to
  models based on weights on literals, and this severely restricts what
  probability distributions can be represented. We develop a measure-theoretic
  perspective on WMC and propose a way to encode conditional weights on literals
  analogously to conditional probabilities. This representation can be as
  succinct as standard WMC with weights on literals but can also expand as
  needed to represent probability distributions with less structure. To
  demonstrate the performance benefits of conditional weights over the addition
  of extra literals, we develop a new WMC encoding for Bayesian networks and
  adapt a state-of-the-art WMC algorithm \textsf{ADDMC} to the new format. Our
  experiments show that the new encoding results in significantly faster
  inference on most benchmark instances. Furthermore, the same idea can be
  adapted to other WMC algorithms and other problem domains.
\end{abstract}

@@

TODO: introduce (also talk more about how Cachet and Ace are used)
\begin{itemize}
\item \texttt{bklm16} \citep{DBLP:conf/ecai/BartKLM16}
\item \textsf{query-dnnf}
  \footnote{\url{http://www.cril.univ-artois.fr/kc/d-DNNF-reasoner.html}}
\item \textsf{DPMC} \citep{DBLP:conf/cp/DudekPV20}
\item tackle a comparable problem, but their encoding was never implemented or
  empirically tested (i.e., the implementation is left as future work)
  \citep{DBLP:conf/scai/VomlelT13}
\end{itemize}

@@

This new measure-theoretic perspective allows us to not only encode any
discrete probability distribution but also improve inference speed. In
\cref{sec:bns}, we demonstrate this by developing a new WMC encoding for
Bayesian networks that uses \emph{conditional weights} on literals (in the
spirit of conditional probabilities) that have literal-based WMC as a special
case. We prove the correctness of the encoding and show how a state-of-the-art
WMC solver \textsf{ADDMC} \citep{DBLP:conf/aaai/DudekPV20} can be adapted to the
new format. \textsf{ADDMC} is a recently-proposed algorithm for WMC based on
manipulating functions on Boolean algebras using an efficient representation for
such functions known as algebraic decision diagrams (ADDs)
\citep{DBLP:journals/fmsd/BaharFGHMPS97}. Finally, in \cref{sec:experiments}, we
show how the new encoding results in faster inference on the vast majority of
benchmark instances, often by one or two orders of magnitude. We explain the
performance benefits by showing how our encoding has asymptotically fewer
variables and ADDs.

@@

Hitherto, four techniques have been proposed for encoding Bayesian networks into
instances of WMC. We will identify them based on the initials of authors as well
as publications years: \texttt{d02} \citep{DBLP:conf/kr/Darwiche02},
\texttt{sbk05} \citep{DBLP:conf/aaai/SangBK05}, \texttt{cd05}
\citep{DBLP:conf/ijcai/ChaviraD05}, and \texttt{cd06}
\citep{DBLP:conf/sat/ChaviraD06}. Below we summarise the observed performance
differences among them. \citet{DBLP:conf/aaai/SangBK05} claim that
\texttt{sbk05} is a smaller encoding than \texttt{d02} with respect to both the
number of clauses and the number of variables but provide no experimental
comparison. \citet{DBLP:conf/ijcai/ChaviraD05} compare \texttt{cd05} with
\texttt{d02} by measuring the time it takes to compile either encoding into an
arithmetic circuit (but do not measure inference time). The more recent encoding
\texttt{cd05} always compiles faster and results in a smaller arithmetic circuit
(as measured by the number of edges). In their subsequent paper, the same
authors perform two sets of experiments (that are relevant to this summary)
\citep{DBLP:conf/sat/ChaviraD06}. First, they compile \texttt{cd05} and
\texttt{cd06} encodings into d-DNNF (i.e., deterministic decomposable negation
normal form \citep{DBLP:journals/jancl/Darwiche01}), measuring both compilation
time and numbers of edges in the d-DNNF diagram. The results are mostly in
favour of \texttt{cd06}. Second, they compare the inference time of
\texttt{sbk05} run with \textsf{Cachet} \citep{DBLP:conf/sat/SangBBKP04} with
the compile times of \texttt{cd05} and \texttt{cd06}, but only on five (types
of) instances. In these experiments, \texttt{cd06} is always faster than
\texttt{cd05}, while the comparison with \texttt{sbk05} is mixed. Sometimes
\texttt{cd06} is orders of magnitude faster than \texttt{sbk05}, sometimes
slightly slower. The performance difference between \texttt{sbk05} and
\texttt{cd05} is even harder to judge: \texttt{sbk05} is better on three out of
five instances and worse on the remaining two. Based on this description, one
would expect \texttt{cd06} to be faster than both \texttt{cd05} and
\texttt{sbk05}, both of which should be faster than \texttt{d02}. The
experiments in \cref{sec:experiments}, however, strongly disagree with this
prediction, showing that the quality of an encoding depends strongly on the
underlying search algorithm or compilation technique.

@@

We also consider the space of all functions from any Boolean algebra to
$\mathbb{R}_{\ge 0}$ together with some operations on those functions. They will
be instrumental in defining WMC as a measure in \cref{sec:wmc_as_measure} and
can be efficiently represented using ADDs. Furthermore, all of the operations
are supported by the CUDD \citep{somenzi1998cudd} package that is used by
\textsf{ADDMC} for ADD manipulation \citep{DBLP:conf/aaai/DudekPV20}. The
definitions of multiplication and projection are as defined by
\citet{DBLP:conf/aaai/DudekPV20}, while others are new.

@@

\Cref{alg:encoding} produces a function with a Boolean algebra as its domain.
This function can be represented by an ADD
\citep{DBLP:journals/fmsd/BaharFGHMPS97}. The core of \textsf{ADDMC} works by
taking an ADD $\psi\colon 2^{U} \to \mathbb{R}_{\ge 0}$ (expressed as a product
of smaller ADDs) and returning $(\exists_U\psi)(\emptyset)$
\citep{DBLP:conf/aaai/DudekPV20}. In this section, we prove that the function
$\phi$ produced by \cref{alg:encoding} can be used by \textsf{ADDMC} to
correctly compute any marginal probability of the Bayesian network that was
encoded as $\phi$.\footnote{Note that it can just as well compute \emph{any}
  probability expressed using the random variables in $\mathcal{V}$.} We begin
with \cref{lemma:cpt} which shows that any conditional weight function produces
the right answer when given a valid encoding of variable-value assignments
relevant to the CPT.

@@

We end with \cref{thm:correctness} that shows how $\phi$ can be combined with an
encoding of a single variable-value assignment so that \textsf{ADDMC} would
compute its marginal probability.

@@

\Cref{alg:encoding} encodes a Bayesian network into a function on a Boolean
algebra, but how does it relate to the standard interpretation of a WMC encoding
as a formula in conjunctive normal form (CNF) together with a collection of
weights? The factors of $\phi$ that restrict the values of indicator variables
for non-binary random variables are already expressed as a product of sums of
0/1-valued functions, i.e., a kind of CNF. Disregarding these functions, each
conditional weight function $\mathrm{CPT}_X$ is represented by a sum with a term
for every subset of $\mathcal{E}^*(X)$. To encode these terms, we introduce
\emph{extended weight clauses} to the WMC format used by \textsf{Cachet}
\citep{DBLP:conf/sat/SangBBKP04}. For instance, here is a representation of the
Bayesian network from \cref{fig:example_bn}:
\[
  \begin{array}{lrrll}
    \lambda\sb{T=l} &\lambda\sb{T=m} &\lambda\sb{T=h} & &0 \\
                    &-\lambda\sb{T=l} &-\lambda\sb{T=m} & &0 \\
                    &-\lambda\sb{T=l} &-\lambda\sb{T=h} & &0 \\
                    &-\lambda\sb{T=m} &-\lambda\sb{T=h} & &0 \\
    w &\lambda\sb{W=1} & &0.5 &0.5 \\
    w &\lambda\sb{F=1} &\lambda\sb{W=1} &0.6 &0.4 \\
    w &\lambda\sb{F=1} &-\lambda\sb{W=1} &0.1 &0.9 \\
    w &\lambda\sb{T=l} &\lambda\sb{W=1} &0.2 &1 \\
    w &\lambda\sb{T=m} &\lambda\sb{W=1} &0.4 &1 \\
    w &\lambda\sb{T=h} &\lambda\sb{W=1} &0.4 &1 \\
    w &\lambda\sb{T=l} &-\lambda\sb{W=1} &0.6 &1 \\
    w &\lambda\sb{T=m} &-\lambda\sb{W=1} &0.3 &1 \\
    w &\lambda\sb{T=h} &-\lambda\sb{W=1} &0.1 &1
  \end{array}
\]
where each indicator variable is eventually replaced with a unique positive
integer. Each line prefixed with a $w$ can be split into four parts: the `main'
variable (always not negated), conditions (possibly none), and two weights. For
example, the line
\[
  \begin{array}{lrrll}
    w &\lambda\sb{T=m} &-\lambda\sb{W=1} &0.3 &1
  \end{array}
\]
encodes the function $0.3[\lambda_{T=m}] \cdot \overline{[\lambda_{W=1}]} +
1\overline{[\lambda_{T=m}]} \cdot \overline{[\lambda_{W=1}]}$ and can be
interpreted as defining two conditional weights: $\nu(T = m \mid W = 0) = 0.3$,
and $\nu(T \ne m \mid W = 0) = 1$, the former of which corresponds to a row in
the CPT of $T$ while the latter is artificially added as part of the encoding.
In our encoding of Bayesian networks, it is always the case that, in each weight
clause, either both weights sum to one, or the second weight is equal to one.
Finally, note that the measure induced by these weights is not probabilistic
(i.e., $\mu(\top) \ne 1$) by itself, but it becomes probabilistic when combined
with the additional clauses that restrict what combinations of indicator
variables can co-occur.

\subsection{Changes to \textsf{ADDMC}}

Here we describe two changes to
\textsf{ADDMC}\footnote{\url{https://github.com/vardigroup/ADDMC}}
\citep{DBLP:conf/aaai/DudekPV20} needed to adapt it to the new format.

First, \textsf{ADDMC} constructs the \emph{Gaifman graph}
of the input CNF formula as an aid for the algorithm's
heuristics. This graph has as vertices the variables of the formula, and there
is an edge between two variables $u$ and $v$ if there is a clause in the formula
that contains both $u$ and $v$. We extend this definition to functions on
Boolean algebras, i.e., the factors of $\phi$. For any pair of distinct
variables $u, v \in U$, we draw an edge between them in the Gaifman graph if
there is a function $\alpha\colon 2^X \to \mathbb{R}_{\ge 0}$ that is a factor
of $\phi$ such that $u \in X$ and $v \in X$. For instance, a factor such as
$\mathrm{CPT}_X$ will enable edges between all distinct pairs of variables in
$\mathcal{E}^*(X)$. Second, even though the function $\phi$ produced by
\cref{alg:encoding} is constructed to have $2^U$ as its domain, sometimes the
domain is effectively reduced to $2^V$ for some $V \subset U$ by the ADD
manipulation algorithms that optimise the ADD representation of a function. For
a simple example, consider $\alpha: 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ defined as
$\alpha(\{a\}) = \alpha(\emptyset) = 0.5$. Then $\alpha$ can be reduced to
$\alpha'\colon 2^{\emptyset} \to \mathbb{R}_{\ge 0}$ defined as
$\alpha'(\emptyset) = 0.5$. To compensate for these reductions, for the original
WMC format with a weight function $w\colon U \cup \{ \neg u \mid u \in U \} \to
\mathbb{R}_{\ge 0}$, \textsf{ADDMC} would multiply its computed answer by
$\prod_{u \in U \setminus V} w(u) + w(\neg u)$. With the new WMC format, we
instead multiply the answer by $2^{|U \setminus V|}$. Each `excluded' variable
$u \in U \setminus V$ satisfies two properties: all weights associated with $u$
are equal to $0.5$ (otherwise the corresponding CPT would depend on $u$, and $u$
would not be excluded), and all other CPTs are independent of $u$ (or they may
have a trivial dependence, where the probability stays the same if $u$ is
replaced with its complement). Thus, the CPT that corresponds to $u$ still
multiplies the weight of every atom in the Boolean algebra by $0.5$, but the
number of atoms under consideration is halved. To correct for this, we multiply
the final answer by two for every $u \in U \setminus V$.

@@

In this section, we describe an experimental study comparing the five WMC
encodings for Bayesian networks when run with \textsf{ADDMC}. The experiments
were run on servers with Intel Xeon Gold 6138 and Intel Xeon E5-2630
processors\footnote{Each instance is run on the same processor for all
  encodings.} running Scientific~Linux~7 with an \SI{8}{\gibi\byte} memory limit
and a \SI{1000}{\second} timeout.

@@

For all encodings other than \texttt{cw}, we use their implementation in
\textsf{Ace}~3.0\footnote{\url{http://reasoning.cs.ucla.edu/ace/}} with
\texttt{-encodeOnly} and \texttt{-noEclause} flags. However, \textsf{Ace} was
not used to encode evidence, as preliminary experiments revealed that the
evidence-encoding implementation contains bugs that can lead to incorrect
answers or a Java exception being thrown on some instances of the data set (and
the source code is not publicly available). Instead, we simply list all the
evidence as additional clauses in the encoding.

Note that both \texttt{cd05} and \texttt{cd06} purposefully produce overly
relaxed encodings that contain extra models and thus yield incorrect
probabilities \citep{DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06}. These
additional models are supposed to be filtered out during circuit compilation
\citep{DBLP:conf/ijcai/ChaviraD05}, but this is not easily achievable with
\textsf{ADDMC}. Nonetheless, we include both encodings in the experiments.

@@

For experimental data, we use the Bayesian networks available with \textsf{Ace}
and
\textsf{Cachet}\footnote{\url{https://www.cs.rochester.edu/u/kautz/Cachet/}},
most of which happen to be binary. We classify them into the following seven
categories:
\begin{itemize*}
\item DQMR and
\item Grid networks as described by \citet{DBLP:conf/aaai/SangBK05},
\item Mastermind, and
\item Random Blocks from the work of \citet{DBLP:journals/ijar/ChaviraDJ06},
\item remaining binary Bayesian networks that include Plan Recognition
  \citep{DBLP:conf/aaai/SangBK05}, Friends and Smokers, Students and Professors
  \citep{DBLP:journals/ijar/ChaviraDJ06}, and \texttt{tcc4f}, and
\item non-binary classic Bayesian networks (\texttt{alarm}, \texttt{diabetes},
  \texttt{hailfinder}, \texttt{mildew}, \texttt{munin1}--\texttt{4},
  \texttt{pathfinder}, \texttt{pigs}, \texttt{water}).
\end{itemize*}
We run \textsf{ADDMC} with each of the five encodings once on each Bayesian
network.

\begin{figure*}[t]
  \centering
  \input{cumulative.tex}%
  \caption{Cumulative number of instances solved by \textsf{ADDMC} over time
    using each encoding.}
  \label{fig:cumulative}
\end{figure*}

\begin{table} % TODO: update or remove
  \centering
  \caption{The numbers of instances (out of 1466) solved by \textsf{ADDMC} with
    each encoding (uniquely, faster than with others, and in total) under the
    described constraints.}%
  \begin{tabular}{lrrr}
    \toprule
    Encoding & Unique & Fastest & Total \\
    \midrule
    \texttt{cd05} & 2 & 3 & 380 \\
    \texttt{cd06} & 0 & 0 & 363 \\
    \texttt{cw} & 234 & 935 & 980 \\
    \texttt{d02} & 40 & 111 & 797 \\
    \texttt{sbk05} & 12 & 44 & 765 \\
    \bottomrule
  \end{tabular}
  \label{tbl:tallies}
\end{table}

\begin{figure*}[t]
  \centering
  \input{scatter.tex}%
  \caption{\textsf{ADDMC} inference time using \texttt{cw} compared to
    \texttt{d02} (left) and \texttt{sbk05} (right) for each instance in all data
    sets.}
  \label{fig:scatter}
\end{figure*}

According to the cumulative plot in \cref{fig:cumulative}, \texttt{cw} is
consistently able to solve more instances with any given time limit than any
other encoding, while the order of the remaining encodings from best to worst is
exactly the opposite of that in previous literature (cf. \cref{sec:related}).
Specifically, \texttt{cw} solves the same number of instances in
\SI{8.144}{\second} as \texttt{d02} does in \SI{1000}{\second}, which is the
second-best encoding in our experiments. Furthermore, \cref{tbl:tallies} shows
that \texttt{cw} was the fastest in 935 out of the 980 (i.e.,
\SI{95.4}{\percent}) instances that it managed to solve. Finally, according to
the scatter plots in \cref{fig:scatter}, \texttt{cw} is particularly promising
on Grid networks while losing to \texttt{d02} on some instances of the
Mastermind data set. We conjecture that this is so either because the particular
structure of the problem confuses the heuristics used by \textsf{ADDMC} or
because these instances allow for significant simplifications that are exploited
by \texttt{d02} but not \texttt{cw}.

@@

\clearpage
\bibliography{paper}

\clearpage
\appendix
\section{Proofs}
